Efficient Thresholded Correlation using Truncated Singular Value Decomposition

### 2017-08-14 threshold/draft.tex
Updated draft following Mike's revision
Switching to the tex version only--no longer using Sweave (draft.tex is the working document now)

%% Instructions for processing this document into a PDF document
%%
%% FIRST NOTE, you need the 'xtable' and 'biclust' packages installed in R.
%% Some of the examples will come from data sets in this package.
%% You can install it from the R console with, for example:
%%
%% install.packages("biclust",repos="http://cran.case.edu")
%% install.packages("xtable",repos="http://cran.case.edu")
%%
%% Now you can prepare the PDF of the paper with any of:
%%
%% From a shell:
%% R CMD Sweave --pdf draft.Rnw
%%
%% From an R prompt:
%% Sweave("draft.Rnw")
%% tools::texi2pdf("draft.tex")
%%
%% From RStudio:
%% click on the little "Compile PDF" button
%%
%% Note that you should probably run the above a few times because
%% LaTeX generally takes a few runs to get reference numbers right.

@@

Because the number of correlation coefficients grows quadratically with the 
number of vectors, simply computing all pairs of correlation coefficients and 
then filtering out coefficients below a given threshold may not be 
computationally feasible for large-scale data applications. As a result, 
considerable attention has been devoted to \emph{pruning methods} that cheaply 
prune away all but the most highly-correlated pairs of vectors.  For example the 
TAPER and related algorithms of Xiong et al. develop pruning rules based on the 
frequency of mutually co-occurring observations (rows) among vector pairs; see 
\cite{prune1, prune2} and the references therein.  Related methods approximate a 
set containing the most highly-correlated pairs using hashing algorithms 
\cite{prune3}. Other methods arrive at approximate sets of the most 
highly-correlated pairs using dimensionality-reduction techniques like the 
discrete Fourier transform \cite{timeseries}.

@@

Let $A=U\Sigma V^T$ be a singular value decomposition of $A$, where
$U\in\mathcal{R}^{m\times k}$, $V\in\mathcal{R}^{n\times k}$,
$U^TU = V^TV = I$,
and $\Sigma\in\mathcal{R}^{k\times k}$ is a diagonal matrix with
diagonal entries $\sigma_1 \ge \sigma_2 \ge \cdots \ge \sigma_k > 0$.  For
later convenience, we write $s_{1:p}=[\sigma_1, \sigma_2, \ldots, \sigma_p]^T$
to be the vector of the first $p\le k$ singular values along the diagonal of
$\Sigma$.  Denote the columns of $V$ as $V=[v_1, v_2, \ldots, v_k]$, where each
$v_j\in\mathcal{R}^n$, $j=1,2,\ldots,k$.  Then $\mathrm{cor}(A) = A^TA =
V\Sigma^2 V^T$ is a symmetric eigenvalue decomposition of the correlation
matrix with nonzero eigenvalues $\sigma_1^2, \sigma_2^2, \ldots, \sigma_k^2$.
The columns of $V$ form an orthonormal basis of the range of the
correlation matrix $A^TA$.

@@

Let $D^i\in\mathcal{R}^{n-i, n}$, $i=1,2,\ldots,n-1$, be the $i$th order finite
difference matrix with $-1$ along the diagonal, $1$ along the $i$th
super-diagonal, and zeros elsewhere.  Then $D^1P v_1$ consists of the
differences of adjacent entries in $P v_1$.  Entries such that $\sigma_1^2 (D^1
P v_1)^2 > 2(1-t)$ correspond to pairs of vectors that do not meet the
correlation threshold, where here and below an exponent applied to a vector
denotes element-wise exponentiation.

Even a 1-d projected interval distance threshold may prune many possible
non-adjacent vector pairs with respect to the ordering defined by $P$ as
illustrated by the example in Figure \ref{fig1}.  However, this observation is
unlikely to rule out many adjacent pairs of vectors in most problems.  For
instance, the maximum distance between adjacent points shown in the example in
Figure \ref{fig1} is less than $\sqrt{2(1-t)}/\sigma_1$, and no pruning of
adjacent pairs of vectors with respect to the ordering $P$ occurs.

Including more terms from Equation \ref{vdist} increases our ability to prune
adjacent pairs of vectors below the correlation threshold with respect to the
ordering defined by $P$.  Including only one more term finds that 96\% of
adjacent vectors fall below the correlation threshold of $0.99$ in the
example shown in Figure \ref{fig1} and are pruned by identifying indices such
that
\[
\sigma_1^2 (D^1 P v_1)^2 +
\sigma_2^2 (D^1 P v_2)^2 > 2(1-t).
\]
Note that we use the permutation $P$ defined by the order of vector $v_1$
throughout.  In general, using $p\le k$ terms to prune pairs of vectors
below the correlation threshold boils down to evaluating the expression
\begin{equation}\label{proj}
(D^1 P V_{1:p} )^2 s_{1:p}^2 > 2(1-t),
\end{equation}
where 
$V_{1:p}$ denotes the first $p$ columns of the
matrix $V$, $s_{1:p}$ the vector of the first $p$ singular values.

@@

\begin{algorithm}\label{prune}
Input: data matrix $A\in \mathcal{R}^{m\times n}$ with columns scaled to have
zero mean and unit norm, correlation threshold $t$,
truncated SVD rank $p<<k$.
\begin{enumerate}
\item Compute the truncated SVD
$AV_{1:p} = U_{1:p}\mathrm{diag}(s_{1:p})$,
where\\
$V_{1:p}=[v_1, v_2, \ldots, v_p]$
and $\mathrm{diag}(s_{1:p})$ denotes the diagonal matrix of the first $p$ singular values.
\item Compute permutation $P$ that orders points in $v_1$.
\item Compute $\ell$, the longest run of successive points in $P v_1$ within the interval $\sqrt{2(1-t)}/\sigma_1$.
\item Compute a set of candidate index pairs $(\tilde{j},\tilde{j}+\tilde{i})$ with respect to the index order defined by the permutation $P$ that possibly meet the correlation threshold
\[
\bigcup_{\tilde{i}=1}^\ell
\left\{
(\tilde{j},\tilde{j}+\tilde{i}) : 
e_{\tilde{j}}^T(D^{\tilde{i}} P V_{1:p} )^2 s_{1:p}^2 \le 2(1-t)
\right\},
\]
where $\tilde{j}=1,2,\ldots,n-\tilde{i}$.
\item If there are judged to be  ``too many'' candidate pairs, increase $p$,
compute $AV_{1:p} = U_{1:p}\mathrm{diag}(s_{1:p})$ and go to step 4.
Continue to increase $p$ in this way until the number of candidate
pairs is sufficiently small or stops decreasing (the best this algorithm can do).
\item Recover the non-permuted candidate column vector
indices $(j,i)$ from $j=e_{\tilde{j}}^T Pq$ and $i=e_{\tilde{j}+\tilde{i}}^TPq$ for every
pair $(\tilde{j},\tilde{j} + \tilde{i})$ in step 4, where
$q\in\mathcal{R}^n = [1,2,\ldots,n]^T$.
\item Compute the full correlation coefficient for each column vector pair identified in
the previous step, applying the correlation threshold to this reduced set of values.
\end{enumerate}
\end{algorithm}
One can cheaply estimate whether there are ``too many'' candidate pairs for a
given dimension $p$ in step 5 by evaluating step 4 for only adjacent pairs of
vectors, $\tilde{i}=1$.

@@

@@

This section presents a number of practical extensions to the main thresholded
correlation idea, most of which are available in the reference R package
implementation. Extensions include finding anti-correlated vectors, finding
thresholded correlations between column vectors from two different matrices,
and application to other distance metrics in place of correlation. All of the
discussed extensions are simple to implement.

@@

\subsection{Spearman correlation}

TODO: WRITE THIS
The idea of evaluating threshold conditions by projection into low dimensional
subspaces applies to distance metrics similarly to correlation, see
\cite{svd-similarity} for a related approach. Lemma \ref{sumlemma} decomposes
Euclidean distance between two vectors in terms of singular basis vectors and
Algorithm \ref{prune} can be applied directly to the problem of finding all
pairs of vectors that meet a thresholded Euclidean distance with small changes
in steps 3, 4, and 7.  By equivalence of norms, thresholded Euclidean distance
can be extended to other distance metrics. The reference \proglang{R}
implementation \cite{sup} includes functions for thresholded distance and
correlation.

\subsection{Most correlated pairs}

TODO: WRITE THIS

The ideas presented here can be adapted to efficiently find the top N most
correlated vector pairs for a given constant N similarly to the work of
Xiong et al. \cite{prune2}.

\section{General matrices and fast truncated SVD}\label{irlba}

For Algorithm \ref{prune} to be used in practice, three challenges need to be 
addressed. First, the algorithm requires an efficient method for computing a 
relatively low-rank truncated SVD of potentially large matrices required by step 
1. Second, the truncated SVD method should be able to cheaply restart to compute 
additional singular vectors as required in step 5 of Algorithm \ref{prune}. 
Third, while the algorithm assumes a standardized matrix, it should work for 
general matrices. Fortunately, each of these challenges is met with a single 
algorithm.

The first two challenges are addressed directly by the augmented implicitly 
restarted Lanczos bidiagonalization algorithm\break(IRLBA) of Baglama and Reichel 
\cite{irlba}. It efficiently computes truncated singular value decompositions of 
either large dense or sparse matrices and the result from the $k$ rank approximati
on can be efficiently used to create an approximation with higher rank. Reference 
implementations are available for \proglang{R} \cite{irlbar}, \proglang{Matlab} 
\cite{irlbam}, and \proglang{Python} \cite{irlbap}.

The third challenge is addressed by relaxing the column mean and scale 
assumptions without introducing significant additional computational or storage
overhead as follows. Assume $A$ is an $m\times n$ real-valued matrix without any 
constant-valued columns. Let $z=[z_1,z_2,\ldots,z_n]\in\mathcal{R}^n$ represent 
the vector of column means of $A$, $w=[w_1^2,w_2^2,\ldots,w_n^2]\in\mathcal{R}^n$
be the vector of squared column norms of $A$, and $W\in\mathcal{R}^{n\times n}$
a diagonal matrix with diagonal entries $1/\sqrt{w_i^2 - m z_i^2}$.  Let
$e\in\mathcal{R}^m$ be a vector of all ones.  Then $(A - ez^T)W$ is a centered
matrix with zero column means and scaled to have unit column norms, and
\[
\mathrm{cor}(A) = W^T (A-ez^T)^T (A-ez^T) W.
\]
%The IRLBA, based on the Lanczos process, is a method of iterated matrix-vector
%products. The main idea behind efficient application of IRLB to correlation
%problems replaces 
Matrix-vector products of the form $Ax$ with $AWx - ez^TWx$ may be used in
the iterations of the algrithm, implicitly working with a scaled and 
centered matrix without explicit standardization. This comes at the cost 
of storing two additional length $n$ vectors and
at most computing two additional vector inner products per matrix-vector product.
The \proglang{R} implementation \cite{irlbar} of IRLBA includes arguments for centering and scaling the input matrix.

@@

Anyone can quickly run the example to experiment with the algorithm and
compare its output with brute force results. The example uses the {\tt
EisenYeast} data from the {\tt biclust} \proglang{R} package \cite{biclust}.
The data consist of 80 experiments involving the
Saccharomyces Cerevisiae yeast measuring gene expression data across 6,221
Affymetrix probes, represented as an $80\times 6221$ real-valued matrix.  The
example finds all pairs of columns (gene expression measurements) that meet or exceed a
Pearson correlation value of 0.95. The choice of projection dimension $p$ affects
the efficiency of the algorithm. For $p=2$ the number of candidate vector pairs
generated in step 4 of Algorithm \ref{prune} is over 3 million. When $p$ is increased
to $10$, the number of candidate vector pairs drops to less than 13,000.
The reference \proglang{R} implementation includes options for
tuning $p$ and relies on the restart ability of the IRLBA to pick up where it left off and
inexpensively extend an existing truncated SVD. The following example uses $p=10$.
\begin{small}
\begin{Schunk}
\begin{Sinput}
> threshold = 0.95
> # Test system values:
> t1=3.13
> t2=1.75
> m1=59.60
> m2=1075.10
> if(Sys.getenv("RUN_EXAMPLES")=="TRUE")
+ {
+ library(tcor)
+ library(doMC)     # local parallel computation
+ library(biclust)  # for the BicatYeast data
+ data(EisenYeast)  # in gene by sample orientation
+ A = t(EisenYeast) # to correlate genes to genes
+ # Register a four-core multicore parallel backend
+ registerDoMC(4)
+ 
+ # Compute using tcor():
+ m1  = sum(gc()[,2])  # Memory in megabytes currently in use
+ t1  = proc.time()
+ tx  = tcor(A, t=threshold, p=10)
+ t1  = (proc.time() - t1)[3] # Time in seconds
+ m1  = sum(gc()[,6]) - m1    # Peak excess memory use during the test
+ 
+ # Fast brute force full correlation:
+ m2  = sum(gc()[,2])
+ t2  = proc.time()
+ mu = colMeans(A)
+ s  = sqrt(apply(A, 2, crossprod) - nrow(A) * mu ^ 2)
+ cs = scale(A, center=mu, scale=s)  # scale and center the data
+ cx = crossprod(cs)                 # full correlation matrix
+ cx  = cx * upper.tri(cx)    # ignore symmetry and diagonal
+ pairs = which(cx >= threshold, arr.ind=TRUE)
+ t2  = (proc.time() - t2)[3]
+ m2  = sum(gc()[,6]) - m2
+ }
\end{Sinput}
\end{Schunk}
\end{small}
Brute-force computation outperforms {\tt tcor()} in this small first example in
speed but consumes substantially more memory.  Each algorithm identifies the
same set of 125 gene expression pairs that exceed the specified correlation
threshold of 0.95, with comparative timing and memory use shown in the
following table.

% latex table generated in R 3.4.1 by xtable 1.8-2 package
% Thu Aug 10 15:41:27 2017
\begin{table}[ht]
\centering
\begingroup\small
\begin{tabular}{lrr}
  \hline
Method & Wall clock time (s) & Peak excess memory use (MB) \\ 
  \hline
tcor & 3.13 & 59.60 \\ 
  brute force & 1.75 & 1075.10 \\ 
   \hline
\end{tabular}
\endgroup
\caption{Gene expression results, threshold=0.95} 
\label{EisenYeast}
\end{table}

@@

% latex table generated in R 3.4.1 by xtable 1.8-2 package
% Thu Aug 10 15:41:27 2017
\begin{table}[ht]
\centering
\begingroup\small
\begin{tabular}{lrrl}
  \hline
Method & Wall clock time (s) & Memory use (MB) &   \\ 
  \hline
tcor (4 core PC) & 10309.00 & 389.10 &  \\ 
  tcor (64 core cluster) & 482.00 & 994.30 & (local) \\ 
   &  & 42282.40 & (remote) \\ 
   \hline
\end{tabular}
\endgroup
\caption{TCGA ACC Methylation450 (80 rows x 394,014 columns, threshold=0.99). Linux cluster results include master R process memory use (local) plus total peak memory used by all 64 worker R processes (remote).} 
\label{TCGA3}
\end{table}From a set of almost 80 billion pairs of vectors, Algorithm \ref{prune} finds 913,601
pairs with correlation values of 0.99 or more on a modest Linux
cluster in about 8 minutes. The same algorithm works, more slowly, in
the constrained memory setting of a desktop PC without modification.

>>>>

@@

Because the number of correlation coefficients grows quadratically with the 
number of vectors, simply computing all pairs of correlation coefficients and 
then filtering out coefficients below a given threshold may not be 
computationally feasible for high-dimensional data. As a result, 
considerable attention has been devoted to \emph{pruning methods} that cheaply 
prune away all but the most highly-correlated pairs of vectors.  For example the 
TAPER and related algorithms of Xiong et al. develop pruning rules based on the 
frequency of mutually co-occurring observations (rows) among vector pairs; see 
\cite{prune1, prune2} and the references therein.  Related methods approximate a 
set containing the most highly-correlated pairs using hashing algorithms 
\cite{prune3}. Other methods arrive at approximate sets of the most 
highly-correlated pairs using dimensionality-reduction techniques like the 
discrete Fourier transform \cite{timeseries}.

@@

Let $A=U\Sigma V^T$ be a singular value decomposition of $A$, where
$U\in\mathcal{R}^{m\times k}$, $V\in\mathcal{R}^{n\times k}$,
$U^TU = V^TV = I$,
and $\Sigma\in\mathcal{R}^{k\times k}$ is a diagonal matrix with
diagonal entries $\sigma_1 \ge \sigma_2 \ge \cdots \ge \sigma_k > 0$.
%%For
%%later convenience, we write $s_{1:p}=[\sigma_1, \sigma_2, \ldots, \sigma_p]^T$
%%to be the vector of the first $p\le k$ singular values along the diagonal of
%%$\Sigma$.
Denote the columns of $V$ as $V=[v_1, v_2, \ldots, v_k]$, where each
$v_j\in\mathcal{R}^n$, $j=1,2,\ldots,k$.  Then $\mathrm{cor}(A) = A^TA =
V\Sigma^2 V^T$ is a symmetric eigenvalue decomposition of the correlation
matrix with nonzero eigenvalues $\sigma_1^2, \sigma_2^2, \ldots, \sigma_k^2$.
The columns of $V$ form an orthonormal basis of the range of the
correlation matrix $A^TA$.

@@

Let $D^i\in\mathcal{R}^{n-i, n}$, $i=1,2,\ldots,n-1$, be the $i$th order finite
difference matrix with $-1$ along the diagonal, $1$ along the $i$th
super-diagonal, and zeros elsewhere.  Then $D^1P v_1$ consists of the
differences of sequential entries in $P v_1$.  Entries such that $\sigma_1^2 (D^1
P v_1)^2 > 2(1-t)$ correspond to pairs of vectors that do not meet the
correlation threshold, where here and below an exponent applied to a vector
denotes element-wise exponentiation.

Even a 1-d projected interval distance threshold may prune many possible
non-sequential vector pairs with respect to the ordering defined by $P$ as
illustrated by the example in Figure \ref{fig1}.  However, this observation is
unlikely to rule out many sequential pairs of vectors in most problems.  For
instance, the maximum distance between sequential points shown in the example in
Figure \ref{fig1} is less than $\sqrt{2(1-t)}/\sigma_1$, and no pruning of
sequential pairs of vectors with respect to the ordering $P$ occurs.

Including more terms from Equation \ref{vdist} increases our ability to prune
sequential pairs of vectors below the correlation threshold with respect to the
ordering defined by $P$.  Including only one more term finds that 96\% of
sequential vectors fall below the correlation threshold of $0.99$ in the
example shown in Figure \ref{fig1} and are pruned by identifying indices such
that
\[
\sigma_1^2 (D^1 P v_1)^2 +
\sigma_2^2 (D^1 P v_2)^2 > 2(1-t).
\]
Note that we use the permutation $P$ defined by the order of vector $v_1$
throughout.  In general, using $p\le k$ terms to prune pairs of vectors
below the correlation threshold boils down to evaluating the expression
\begin{equation}\label{proj}
(D^1 P V_{1:p} )^2 s_{1:p}^2 > 2(1-t),
\end{equation}
where 
$V_{1:p}$ denotes the first $p$ columns of the
matrix $V$, $s_{1:p}$ the vector of the first $p$ singular values.

@@

\begin{algorithm}\label{prune}
Input: data matrix $A\in \mathcal{R}^{m\times n}$ with columns scaled to have
zero mean and unit norm, correlation threshold $t$,
truncated SVD rank $p<<k$.
\begin{enumerate}
\item Compute the truncated SVD
$AV_{1:p} = U_{1:p}\mathrm{diag}(s_{1:p})$,
where\\
$V_{1:p}=[v_1, v_2, \ldots, v_p]$
and $\mathrm{diag}(s_{1:p})$ denotes the diagonal matrix of the first $p$ singular values.
\item Compute permutation $P$ that orders points in $v_1$.
\item Compute $\ell$, the longest run of successive points in $P v_1$ within the interval $\sqrt{2(1-t)}/\sigma_1$.
\item Compute a set of candidate index pairs $(\tilde{j},\tilde{j}+\tilde{i})$ with respect to the index order defined by the permutation $P$ that possibly meet the correlation threshold
\[
\bigcup_{\tilde{i}=1}^\ell
\left\{
(\tilde{j},\tilde{j}+\tilde{i}) : 
e_{\tilde{j}}^T(D^{\tilde{i}} P V_{1:p} )^2 s_{1:p}^2 \le 2(1-t)
\right\},
\]
where $\tilde{j}=1,2,\ldots,n-\tilde{i}$.
\item If there are judged to be  too many candidate pairs, increase $p$,
compute $AV_{1:p} = U_{1:p}\mathrm{diag}(s_{1:p})$ and go to step 4.
Continue to increase $p$ in this way until the number of candidate
pairs is sufficiently small or stops decreasing (the best this algorithm can do).
\item Recover the non-permuted candidate column vector
indices $(j,i)$ from $j=e_{\tilde{j}}^T Pq$ and $i=e_{\tilde{j}+\tilde{i}}^TPq$ for every
pair $(\tilde{j},\tilde{j} + \tilde{i})$ in step 4, where
$q\in\mathcal{R}^n = [1,2,\ldots,n]^T$.
\item Compute the full correlation coefficient for each column vector pair identified in
the previous step, applying the correlation threshold to this reduced set of values.
\end{enumerate}
\end{algorithm}
One can cheaply estimate whether there are too many candidate pairs for a
given dimension $p$ in step 5 by evaluating step 4 for only sequential pairs of
vectors, $\tilde{i}=1$.

@@

\subsection{General matrices and fast truncated SVD}\label{irlba}

Three practical challenges need to be addressed for Algorithm \ref{prune} to be
used in practice:
\begin{enumerate}
\item The algorithm needs a computationally efficient way to compute
a low-rank truncated SVD of potentially large sparse or dense matrices.
\item The truncated SVD method should be able to cheaply restart to
compute additional singular vectors as needed by step 5 of Algorithm
\ref{prune}.
\item Algorithm \ref{prune} as stated assumes a standardized matrix;
it needs to be extended to efficiently apply to general matrices.
\end{enumerate}
%For Algorithm \ref{prune} to be used in practice, three challenges need to be 
%addressed. First, the algorithm requires an efficient method for computing a 
%relatively low-rank truncated SVD of potentially large matrices required by step 
%1. Second, the truncated SVD method should be able to cheaply restart to compute 
%additional singular vectors as required in step 5 of Algorithm \ref{prune}. 
%Third, while the algorithm assumes a standardized matrix, it should work for 
%general matrices. Fortunately, each of these challenges is met with a single 
%algorithm.

The first two challenges are addressed directly by the augmented implicitly
restarted Lanczos bidiagonalization algorithm (IRLBA) of Baglama and Reichel
\cite{irlba}. It efficiently computes truncated singular value decompositions
of either large dense or sparse matrices and the result from an existing rank
$k$ approximation can be efficiently extended to higher-rank approximations by
restarting.  Reference implementations are available for \proglang{R}
\cite{irlbar}, \proglang{Matlab} \cite{irlbam}, and \proglang{Python}
\cite{irlbap}.

The third challenge is addressed by relaxing the column mean and scale
assumptions without introducing significant additional computational or storage
overhead as follows. Assume $A$ is an $m\times n$ real-valued matrix without
any constant-valued columns. Let $z=[z_1,z_2,\ldots,z_n]\in\mathcal{R}^n$
represent the vector of column means of $A$,
$w=[w_1^2,w_2^2,\ldots,w_n^2]\in\mathcal{R}^n$ be the vector of squared column
norms of $A$, and $W\in\mathcal{R}^{n\times n}$ a diagonal matrix with diagonal
entries $1/\sqrt{w_i^2 - m z_i^2}$.  Let $e\in\mathcal{R}^m$ be a vector of all
ones.  Then $(A - ez^T)W$ is a centered matrix with zero column means and
scaled to have unit column norms, and
\[
\mathrm{cor}(A) = W^T (A-ez^T)^T (A-ez^T) W.
\]
%The IRLBA, based on the Lanczos process, is a method of iterated matrix-vector
%products.
IRLBA does not directly require the matrix $A$, but instead only needs to
compute matrix-vector products with it.  The main idea behind efficient
application of IRLBA to correlation problems replaces matrix-vector products of
the form $Ax$ with $AWx - ez^TWx$, implicitly working with a scaled and
centered matrix without explicit standardization. This comes at the cost of
storing two additional length $n$ vectors and at most computing two additional
vector inner products per matrix-vector product.  The \proglang{R}
implementation \cite{irlbar} of IRLBA includes arguments for centering and
scaling the input matrix.

@@

A number of practical extensions apply to the main thresholded correlation
idea, most of which are available in the reference R package implementation.
Extensions include finding anti-correlated vectors, finding thresholded
pairwise correlations between column vectors from two different matrices,
application to other distance metrics in place of correlation, and, similarly
to the work of Xiong et al. \cite{prune2}, efficiently finding the top N most
correlated vector pairs for a given constant N.

All of the discussed extensions are simple to implement. This section
outlines key ideas behind the extensions.

\subsection{Spearman correlation and application to other distance metrics}

The idea of evaluating threshold conditions by projection into low dimensional
subspaces applies directly to the Euclidean distance metric, see
\cite{svd-similarity} for a related approach. Lemma \ref{sumlemma} decomposes
Euclidean distance between two vectors in terms of singular basis vectors and
Algorithm \ref{prune} can be applied directly to the problem of finding all
pairs of vectors that meet a thresholded Euclidean distance with small changes
in steps 3, 4, and 7.  By equivalence of norms, thresholded Euclidean distance
can be extended to other distance metrics with increased computational cost.
The reference \proglang{R} implementation \cite{sup} includes functions for
thresholded distance and correlation.

Spearman rank correlation may be found by first transforming the data to ranked
values followed by thresholded correlation using Algorithm \ref{prune} on the
ranked values.

@@

Very similarly, we can compute pairwise thresholded correlation between pairs
of column vectors of two distinct matrices by a truncated SVD of a matrix
formed by coulmn-wise concatenation of the two matrices. The resulting
thresholded pairs are then grouped by original column location to consider only
pairs consisting of column vectors from each original matrix.

\subsection{Top $N$ most correlated pairs}

A procedure that finds the top $N$ most correlated pairs for a fixed integer
$N$ proceeds iteratively starting with a small initial value of $p$, for
instance $p = 0.999$.  Recall that the maximum number of pairs meeting a
threshold is available by step 4 of Algorithm \ref{prune}.  Then:

\begin{enumerate}
\item[a.] Reduce $p$ and repeat steps 1 -- 4 of Algorithm \ref{prune} until
the maximum number of pairs meething the threshold $p$ is greater than $N$.
\item[b.] Complete steps 5 -- 7 of Algorithm \ref{prune}. If the number of
pairs meeting the threshold is less than $N$, reduce $p$ and goto step a.
\end{enumerate}

The procedure then relies on the ability to cheaply restart IRLBA to expand an
existing truncated SVD.

@@

Anyone can quickly run the example to experiment with the algorithm and
compare its output with brute force results. The example uses the {\tt
EisenYeast} data from the {\tt biclust} \proglang{R} package \cite{biclust}.
The data consist of 80 experiments involving the
Saccharomyces Cerevisiae yeast measuring gene expression data across 6,221
Affymetrix probes, represented as an $80\times 6221$ real-valued matrix.  The
example finds all pairs of columns (gene expression measurements) that meet or exceed a
Pearson correlation value of 0.95. The choice of projection dimension $p$ affects
the efficiency of the algorithm. For $p=2$ the number of candidate vector pairs
generated in step 4 of Algorithm \ref{prune} is over 3 million. When $p$ is increased
to $10$, the number of candidate vector pairs drops to less than 13,000.
The reference \proglang{R} implementation includes options for
tuning $p$ and relies on the restart ability of the IRLBA to pick up where it left off and
inexpensively extend an existing truncated SVD. The following example uses $p=10$.
\begin{small}
\begin{Schunk}
\begin{Sinput}
threshold = 0.95
library(tcor)
library(doMC)     # local parallel computation
library(biclust)  # for the BicatYeast data
data(EisenYeast)  # in gene by sample orientation
A = t(EisenYeast) # to correlate genes to genes
# Register a four-core multicore parallel backend
registerDoMC(4)

# Compute using tcor():
m1  = sum(gc()[, 2])  # Memory in megabytes currently in use
t1  = proc.time()
tx  = tcor(A, t=threshold, p=10)
t1  = (proc.time() - t1)[3] # Time in seconds
m1  = sum(gc()[, 6]) - m1    # Peak excess memory use during the test

# Fast brute force full correlation:
m2  = sum(gc()[, 2])
t2  = proc.time()
mu = colMeans(A)
s  = sqrt(apply(A, 2, crossprod) - nrow(A) * mu ^ 2)
cs = scale(A, center=mu, scale=s)  # scale and center the data
cx = crossprod(cs)                 # full correlation matrix
cx  = cx * upper.tri(cx)    # ignore symmetry and diagonal
pairs = which(cx >= threshold, arr.ind=TRUE)
t2  = (proc.time() - t2)[3]
m2  = sum(gc()[, 6]) - m2
\end{Sinput}
\end{Schunk}
\end{small}
Brute-force computation outperforms {\tt tcor()} in this small first example in
speed but consumes substantially more memory.  Each algorithm identifies the
same set of 125 gene expression pairs that exceed the specified correlation
threshold of 0.95, with comparative timing and memory use shown in the
following table.

% latex table generated in R 3.4.0 by xtable 1.8-2 package
% Sat Aug 12 09:35:30 2017
\begin{table}[ht]
\centering
\begingroup\small
\begin{tabular}{lrr}
  \hline
Method & Wall clock time (s) & Peak excess memory use (MB) \\ 
  \hline
tcor & 3.13 & 59.60 \\ 
  brute force & 1.75 & 1075.10 \\ 
   \hline
\end{tabular}
\endgroup
\caption{Gene expression results, threshold=0.95} 
\label{EisenYeast}
\end{table}

@@

% latex table generated in R 3.4.0 by xtable 1.8-2 package
% Sat Aug 12 09:35:32 2017
\begin{table}[ht]
\centering
\begingroup\small
\begin{tabular}{lrrl}
  \hline
Method & Wall clock time (s) & Memory use (MB) &   \\ 
  \hline
tcor (4 core PC) & 10309.00 & 389.10 &  \\ 
  tcor (64 core cluster) & 482.00 & 994.30 & (local) \\ 
   &  & 42282.40 & (remote) \\ 
   \hline
\end{tabular}
\endgroup
\caption{TCGA ACC Methylation450 (80 rows x 394,014 columns, threshold=0.99). Linux cluster results include master R process memory use (local) plus total peak memory used by all 64 worker R processes (remote).} 
\label{TCGA3}
\end{table}From a set of almost 80 billion pairs of vectors, Algorithm \ref{prune} finds 913,601
pairs with correlation values of 0.99 or more on a modest Linux
cluster in about 8 minutes. The same algorithm works, more slowly, in
the constrained memory setting of a desktop PC without modification.
